Linting Linearity in Core/System FC

### 2023-01-11 report.tex
Start linear lambda calculus grammar

\usepackage{todonotes}
\usepackage{cmll}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

@@

However, currently, typechecking linearity in Core is broken in the face of
optimizations! The reason is not evidently clear: if we can typecheck linearity
in the surface level Haskell why can't we in Core? In fact, if we were to
typecheck the programs in Core as they were at the surface level, we would
succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
transformations) and hence end up different from how they were originally
written. For example, the following recursive let definition, where $x$ is a
linear variable that must be used exactly once, would not typecheck in a source
linear Haskell program since the typechecker cannot tell that $x$ is used
linearly, but this program might occur naturally in Core from
transformations on a program that did succesfully typecheck:
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

Is this program really still linear? If one looks at it

@@

Despite \emph{want} a linearly-typed core
because a linearly-typed Core ensures that desugaring
Haskell and optimizing transformations don't destroy linearity.

@@

% Linear type systems add expressiven at the type level regarding values that
% must be used exactly once thorought their lifetime.

@@

@@

A linear type system is a type system in which ...!

@@

>>>>

\usepackage{todonotes}
\usepackage{cmll}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathpartir}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyvrb}

% BNF grammars
\usepackage[center]{formal-grammar}

\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small}
\DefineVerbatimEnvironment{example}{Verbatim}{fontsize=\small}

@@

However, currently, typechecking linearity in Core is broken in the face of
optimizations! The reason is not evidently clear: if we can typecheck linearity
in the surface level Haskell why can't we in Core? In fact, if we were to
typecheck the programs in Core as they were at the surface level, we would
succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
transformations) and hence end up different from how they were originally
written. For example, the following recursive let definition, where $x$ is a
linear variable that must be used exactly once, would not typecheck in a source
Haskell program since the typechecker cannot tell that $x$ is used linearly, but
this program might occur naturally in Core from transformations on a program
that did succesfully typecheck:
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

Is this program really still linear? Yes, but ...

% If one looks at it

@@

% Despite \emph{want} a linearly-typed core
% because a linearly-typed Core ensures that desugaring
% Haskell and optimizing transformations don't destroy linearity.

@@

@@

% TODO? Another defining feature of Haskell is its powerful type system. The
% interaction between the new linearity polymorphism and the existing
% type-level facilities is quite interesting

@@

A linear type system is a type system in which linear resources must be used
exactly once...!

@@

\begin{grammar}
    \firstcase{\t}{\star}{Unit type}
    \otherform{t \tensor t'}{Multiplicative conjunction type}
    % t ::= a $\lolli$ b
\end{grammar}
